Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-12 paper/main.tex
done with section 3

\begin{figure*}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{7cm}
    \begin{center}
   \begin{lstlisting}[]
void c_down (int x,int y) {
  if (x>y) { tick(1); c_up(x-1,y); }
}
void c_up (int x, int y) {
  if (y+1<x) {tick(1); c_down(x,y+2);}
}
   \end{lstlisting}
\vspace{-2.5ex}
$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
\\[.4\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5cm}
    \begin{center}
   \begin{lstlisting}
for (; l >= 8; l -= 8)
  // process one block
  tick(N);
for (; l > 0; l--)
  // save leftovers
  tick(1);
   \end{lstlisting}
\vspace{-2.5ex}
$XXX$
\\[.4\baselineskip]
      {\bf t61}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5.5cm}
    \begin{center}
   \begin{lstlisting}

   \end{lstlisting}
\vspace{-2.5ex}
\begin{lstlisting}
for (;;) {
  do { l++; tick(1); } 
    while (l < h && *);
  do { h--; tick(1); } 
    while (h > l && *);
  if (h <= l) break;
  tick(1); /* swap elems. */ }
\end{lstlisting}
$XXX$
\\[.4\baselineskip]
      {\bf t62}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Exampel \emph{t39} shows two mutually-recursive functions
  with the computed tick bounds.  The derived constant factors are
  tight.}
  \label{fig:ex_list_2}
\end{figure*}

\paraskip
\paragraph{Examples: Challenging Loops.}

@@

@@

@@

@@

\paraskip
\paragraph{Examples: Compositionality.}

With two concrete examples from open-source projects we demonstrate
that the compositionality of our method is indeed crucial in practice.

The following example is typical in cryptographic primitives
implementations.  These primitives work by block, so when data of
arbitrary length needs to be processed, it is munched in blocks and
the leftovers are stored in a buffer for future use when more data
is available.  This looping pattern is particularly well handled by
our tool because the LP solver is leveraged to generate a tight bound.
The optimization problem it is given encodes that constant bounds are
preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
\toolname is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
counted as constant.
%
\jan{N is unbound above. The explanation must be much slower
  here. Explain what the code does and how the ticks account for the
  cost. Then: What is the resource bound?. Then: What do we derive?
  Btw: Why do you use tick(1) and not tick(M)?}
%
\begin{lstlisting}
for (; l >= 8; l -= 8)
    // process one block
    tick(N);
for (; l > 0; l--)
    // save leftovers
    tick(1);
\end{lstlisting}
%
It is worth taking some time to explain the constant contribution
to the resource bound when $N<8$.  Remark first that the cost of
the second loop is $|[0,l]|$, and after the first one, we still have
$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
potential units instead.  This explains the somewhat intriguing value
of the constant term in the bound.  All this reasoning is done within
our tool automatically and simply relies on general rules presented
in further sections.

This level of precision and compositionality is only achieved by
our work, no other available tool will attain the above tight bounds
on this common kind of looping code.  Thanks to the modularity of
amortized analysis the intricate interactions between the bounds of
the two loops can be modeled as one linear program, and because LP
solvers can optimize the quality of the final bound we can keep
it close to the actual resource consumption.

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && *);
    do { h--; tick(1); } while (h > l && *);
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
prove the termination of the loop and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classic technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively, resulting in a loose quadratic bound or no bound at all.

@@

\paragraph{Judgements.}

>>>>

\begin{figure*}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{7cm}
    \begin{center}
   \begin{lstlisting}[]
void c_down (int x,int y) {
  if (x>y) { tick(1); c_up(x-1,y); }
}
void c_up (int x, int y) {
  if (y+1<x) {tick(1); c_down(x,y+2);}
}
   \end{lstlisting}
\vspace{-2.5ex}
$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
\\[.4\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5cm}
    \begin{center}
   \begin{lstlisting}
for (; l >= 8; l -= 8)
  // process one block
  tick(N);
for (; l > 0; l--)
  // save leftovers
  tick(1);
   \end{lstlisting}
\vspace{-2.5ex}
$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
\\[.4\baselineskip]
      {\bf t61}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5.5cm}
    \begin{center}
   \begin{lstlisting}

   \end{lstlisting}
\vspace{-2.5ex}
\begin{lstlisting}
for (;;) {
  do { l++; tick(1); } 
    while (l < h && *);
  do { h--; tick(1); } 
    while (h > l && *);
  if (h <= l) break;
  tick(1); /* swap elems. */ }
\end{lstlisting}
$2+3|[l,h]|$
\\[.4\baselineskip]
      {\bf t62}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Example \emph{t39} shows two mutually-recursive functions
  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
  demonstrate the unique compositionality of our system. In \emph{61},
  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
  are tight.}
  \label{fig:ex_list_2}
\end{figure*}

\paraskip
\paragraph{Challenging Loops.}

@@

\paraskip
\paragraph{Nested and Sequenced Loops.}

@@

\paraskip
\paragraph{Mutual Recursive Functions.}

@@

\paraskip
\paragraph{Compositionality.}

With two concrete examples from open-source projects we demonstrate
that the compositionality of our method is indeed crucial in practice.
\jan{what projects?}

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations of block-based
cryptographic primitives: Data of arbitrary length, is consumed in
blocks and the leftover is stored in a buffer for future use when more
data is available.  This looping pattern is particularly well handled
by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the
cost in the second loop (\code{tick(1)}) are random choices and
$\toolname$ would also derive tight bound for other values.

To understand the resource bound for the case $N<8$, first note that the cost of
the second loop is $|[0,l]|$.  After the first loop, we still have
$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
potential units instead.  
%
This level of precision and compositionality is only achieved by
our novel analysis, no other available tool derives the aforementioned
 tight bounds.

Example \emph{t62} (\pref{fig:ex_list_2}) 
is the inner loop of a quick sort implementation in cBench.  More
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|.$  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
find a bound and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classical technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively.

@@

@@

\paraskip
\paragraph{Judgements.}
